Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-06-05 src/tex/ms.tex
fixed the majority of the yellow comments, but I need to add in some discussion to the pr

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO Virgo Kagra Gravitational Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

Traditionally, population synthesis studies simulate merging DCO populations
with a Monte-Carlo approach.  Using theoretically motivated distributions of
initial parameters like age, metallicity, component mass, orbital separation,
and eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. This approach relies on several assumptions, both in
the simulations and model parameterizations fitted to the GW data, which are
unlikely to be fully correct.  For example, there is no reason to suppose that
the entire population of DCOs should evolve with the same fixed hyperparameters;
there is no a priori reason the fitted parameterized model should capture the
relevant features of the observed DCO distributions to correspond to the binary
physics of interest; etc. A better approach would be to compare progenitor
parameters and hyperparameter settings directly to DCO observations, treating
the population synthesis model as a mapping from progenitor to merger
parameters; merger parameters can then be mapped into observations using a
gravitational waveform family, and parameter inference proceeds in the usual
way, propagating information back up the mapping chain
\citep[e.g.][]{Veitch2015}.  A similar methodology, though with \emph{fixed}
hyperparameter settings, was advanced in \citet{Andrews2018,Andrews2021}, as we
discuss in more detail below.

@@

We apply the method described in section \ref{sec:method} to the first GW event GW150914.
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space in both the
root-finding stage and the MCMC stage. For progenitor parameters, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep{Broekgaarden2021}.
For hyperparameters, we choose parameters that affect
binary evolution in \texttt{COSMIC} most significantly for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$.

@@

For high-mass binaries, \texttt{COSMIC} struggles to produce events above the
pair instability supernova mass cutoff $m_{1,{\rm GW}}$, so the posterior in the
evolutionary parameter space only corresponds to part of the posterior in the
observable space below this cutoff, and therefore events beyond the PISN mass
gap also have higher KL divergence.

>>>>

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

Traditionally, population synthesis studies simulate merging DCO populations
with a Monte-Carlo approach.  Using theoretically motivated distributions of
initial parameters like age, metallicity, component mass, orbital separation,
and eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings like the stability 
of Roche-overflow mass transfer, common envelope ejection efficiency, or the 
strength of compact-object natal kicks, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. This approach relies on several assumptions, both in
the simulations and model parameterizations fitted to the GW data, which are
unlikely to be fully correct.  For example, there is no reason to suppose that
the entire population of DCOs should evolve with the same fixed hyperparameters;
there is no a priori reason the fitted parameterized model should capture the
relevant features of the observed DCO distributions to correspond to the binary
physics of interest; etc. A better approach would be to compare progenitor
parameters and hyperparameter settings directly to DCO observations, treating
the population synthesis model as a mapping from progenitor to merger
parameters; merger parameters can then be mapped into observations using a
gravitational waveform family, and parameter inference proceeds in the usual
way, propagating information back up the mapping chain
\citep[e.g.][]{Veitch2015}.  A similar methodology, though with \emph{fixed}
hyperparameter settings, was advanced in \citet{Andrews2018,Andrews2021}, as we
discuss in more detail below.

@@

We apply the method described in section \ref{sec:method} to the first GW event GW150914 \citep{GW150914}.
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space in both the
root-finding stage and the MCMC stage. For progenitor parameters, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.
For hyperparameters, we choose parameters that affect
binary evolution in \texttt{COSMIC} most significantly for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$.

@@

For high-mass binaries, \texttt{COSMIC} struggles to produce events above the
pair instability supernova mass cutoff, so the posterior in the
evolutionary parameter space only corresponds to part of the posterior in the
observable space below this cutoff, and therefore events beyond the lower edge of the PISN mass
gap also have higher KL divergence.
